From tides to tails: \\ Does the Milky Way affect Sculptor and Ursa Minor?

### 2025-03-21 thesis/introduction.tex
thesis updates

  \begin{itemize}
  \tightlist
  \item
    EN21, peñarrubia+09, etc.
  \end{itemize}
\item
  Instances of dwarfs undergoing weird processess
\item
  Alternative processes and uncertainties in the evolution of dwarfs
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  sestito+23a
\item
  tolstoy+22
\item
  battaglia+09
\item
  arroyo-polonio+22
\end{itemize}

@@

\begin{figure}
\centering
\includegraphics{figures/scl_selection.png}
\caption{Sculptor selection criteria. This will be a super long caption
to test the figure list.}\label{fig:sculptor_selection}
\end{figure}

@@

@@

Note that if we remove the spatial component of the likelihood, then
\(f_{\rm sat}\) represents a global normalization

@@

\begin{figure}
\centering
\includegraphics{figures/scl_tidal_tails.png}
\caption{Density profiles}\label{fig:sculptor_tidal_tails}
\end{figure}

\begin{itemize}
\item
  The above figure shows the distribution of member stars (orange), PM
  \& CMD only selected stars (blue) and all stars (passing quality cuts,
  black)
\item
  There are no apparent overdensities in the PM \& CMD only selected
  stars to suggest the presence of a tidal tail
\item
  This means that at least at the level of where the background density
  dominates, we can exclude models which produce tidal tails brighter
  than a density of
  \(\Sigma_\star \approx 10^{-2}\,\text{Gaia-stars\ arcmin}^{-2} \approx 10^{-6} \, {\rm M_\odot\ kpc^{-2}}\)
  (TODO assuming a distance of \ldots{} and stellar mass of \ldots).
\end{itemize}

@@

\begin{figure}
\centering
\includegraphics{figures/scl_density_methods.png}
\caption[Density profiles]{add simple cuts \& delve to convince
more\ldots{}}\label{fig:sculptor_observed_profiles}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313130050775.png}
\caption{image-20250313130050775}
\end{figure}

\begin{figure}
\centering
\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313130110550.png}
\caption{image-20250313130110550}
\end{figure}

\begin{figure}
\centering
\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313130043114.png}
\caption{image-20250313130043114}
\end{figure}

@@

\subsection{Summary}\label{summary}

>>>>

  \begin{itemize}
  \tightlist
  \item
    \citet{EN2021}, \citet{PNM2008}, etc.
  \end{itemize}
\item
  Instances of dwarfs undergoing weird processess
\item
  Alternative processes and uncertainties in the evolution of dwarfs
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  \citet{sestito+2023a}
\item
  \citet{tolstoy+2023}, \citet{arroyo-polonio+2023},
  \citet{arroyo-polonio+2024}
\item
  \citet{battaglia+2008}
\item
  \citet{iorio+2019}
\end{itemize}

@@

\begin{figure}
\centering
\includegraphics{figures/scl_selection.pdf}
\caption[Sculptor selection criteria]{The selection criteria for Scl
members. Members are red and all field stars (satisfying quality
criteria) are in light grey. \textbf{Top left:} tangent plane.
\textbf{Top Right:} Colour magnitude
diagram.}\label{fig:sculptor_selection}
\end{figure}

@@

Note that if we remove the spatial component of the likelihood, then
\(f_{\rm sat}\) represents a global normalization.

@@

@@

\begin{figure}
\centering
\includegraphics{figures/scl_tidal_tails.pdf}
\caption[Tidal tails]{The distribution of member stars (orange), PM \&
CMD only selected stars (blue) and all stars (passing quality cuts,
black).}\label{fig:sculptor_tidal_tails}
\end{figure}

\begin{itemize}
\item
  There are no apparent overdensities in the PM \& CMD only selected
  stars to suggest the presence of a tidal tail
\item
  This means that at least at the level of where the background density
  dominates, we can exclude models which produce tidal tails brighter
  than a density of
  \(\Sigma_\star \approx 10^{-2}\,\text{Gaia-stars\ arcmin}^{-2} \approx 10^{-6} \, {\rm M_\odot\ kpc^{-2}}\)
  (TODO assuming a distance of \ldots{} and stellar mass of \ldots).
\end{itemize}

@@

\begin{figure}
\centering
\includegraphics{figures/scl_density_methods.pdf}
\caption{Density profiles}\label{fig:sculptor_observed_profiles}
\end{figure}

Figure:

@@

\begin{figure}
\centering
\includegraphics{figures/classical_dwarf_profiles.pdf}
\caption[Sculptor and UMi versus classical dwarfs]{Density profiles for
each dwarf galaxy.}
\end{figure}

\begin{table*}[t]
\centering
\caption{Sérsic profile fits and basic structural parameters for classical and bright dwarfs around the MW.}
\label{tbl:Sérsic-profile-fits-and-basic-structural-parameters-for-classical-and-bright-dwarfs-around-the-MW}
\begin{tabular}{llllll}
\toprule
galaxy & Ell. & Sersic Rh & sersic n & Sersic mass & \\
\midrule
Fornax &  &  &  &  & \\
Leo I &  &  &  &  & \\
Sculptor &  &  &  &  & \\
Leo II &  &  &  &  & \\
Carina &  &  &  &  & \\
Sextans I &  &  &  &  & \\
Ursa Minor &  &  &  &  & \\
Draco &  &  &  &  & \\
Canes Venatici I &  &  &  &  & \\
Crater II &  &  &  &  & \\
Bootes I &  &  &  &  & \\
Bootes III &  &  &  &  & \\
Antlia II &  &  &  &  & \\
\bottomrule
\end{tabular}
\end{table*}

Cannot resolve

\begin{itemize}
\tightlist
\item
  reticulum II yyy
\end{itemize}

@@

\section{Simulation-based
motivations}\label{simulation-based-motivations}

To motivate why a tidal interaction may give rise to the observed
density profiles, we create a toy simulation following \citet{PNM2008}.

\begin{itemize}
\item
  NFW initial conditions (sculptor like, vcm, rcm)
\item
  Evolved in x-y plane using \citet{EP2020} potential for
  \textasciitilde{} 5Gyr with pericentre of 15 kpc and apocentre of 100
  kpc.
\item
  Exponential initial stellar profile.
\end{itemize}

As a dark matter halo is perturbed on a pericentric passage with the
milky way,

\begin{itemize}
\tightlist
\item
  Tidal stress heats halo slightly
\item
  Mass loss, particularly of loosely bound particles
\end{itemize}

The stellar component tracers will similarly follow the behaviour of the
dark matter.

An emperical estimate of where the simulation's stars are becoming
unbound is, as stated in \citet{PNM2008}, the break radius \[
R_b = C\,\sigma_{v}\,\Delta t
\] where \(\sigma_v\) is the present line of sight velocity dispersion ,
\(\delta t\) is the time since pericentre, and \(C \approx 0.55\) is a
fit. The idea motivating this equation is stars in the inner regions
will have dynamically equilibriated to the new potential (phase mixed),
however the outer regions are no longer in steady state, so we have to
wait until the crossing time reaches them as well.

As illustrated in fig.~\ref{fig:toy_profiles}, the density profile
initially stars off exponential. At increasing times since the first
pericentric passage, the break radius, appearing as an apparent
separation between the slopes of the inner and outer profile, increases.

\begin{figure}
\centering
\includegraphics{figures/scl_umi_vs_idealized.pdf}
\caption[Idealized simulations match Scl and UMi]{Sculptor and UMi's
profiles are well-matched to an idealized
simulation}\label{fig:toy_profiles}
\end{figure}

\begin{figure}
\centering
\includegraphics{figures/idealized_break_radius.pdf}
\caption[Break radius validation]{The break radius of the simulations is
set by}\label{fig:idealized_break_radius}
\end{figure}

From this argument, we note that the following properties must be
approximately true for tides to occur:

\begin{itemize}
\tightlist
\item
  Close enough pericentre. The other break radius \(r_J\) implies that
  if the host density is 3x the satellite, stars will be lost
\item
  Corresponding time since last pericentre: If the time since last
  pericentre is not \(\sim\) consistent with an observed break in the
  density profile, then tides
\item
  Halo evolution. As found in \citet{EN2021}, galaxies evolve along well
  defined tidal tracks (assuming spherical, isotropic, NFW halo, which
  may not be true, see \ldots). These tracks tend to ``puff up'' the
  stellar component while also removing dark matter mass, leaving a
  smaller, compacter DM halo with a more extended stellar component.

  \begin{itemize}
  \tightlist
  \item
    This information is mostly related to the statistical initial
    distribution of satellites from cosmology {[}ludlow+2016;
    \citet{fattahi+2018}{]}
  \end{itemize}
\end{itemize}

\section{Summary}\label{summary}

### 2025-03-21 thesis/main.tex
thesis updates

\title{Tidal to tell: \\
    Does the Milky Way affect Sculptor and Ursa Minor?
}
\author{Daniel A. Boyea}

@@

\makeatletter
\begin{titlepage}
   \begin{center}
       \textbf{\@title}\\
        \skipline

@@

    in the Department of Physics and Astronomy\\
       \vfill
       ©\@author, 2025\\
       University of Victoria\\
       \vspace*{3\baselineskip}

       All rights reserved. This thesis may not be reproduced in whole or in part,
by photocopy or other means, without the permission of the author. \\
\skipline
   \end{center}
We acknowledge and respect the \Lekwnen{} (Songhees and X\textsuperscript{w}seps\schwa{}m/
Esquimalt) Peoples on whose territory the university stands, and the
\Lekwnen{} and ${\underaccent{\bar}{\rm W}}$S\'ANE\'C Peoples whose historical relationships with the
land continue to this day. 
\end{titlepage}

@@

% Abstract of the paper
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
% context
Scl and UMi are weird bois.

-\rule[2.4pt]{3pt}{0.75pt}
I\rule{1.2pt}{8pt}
{\bf I}\rule{2.0pt}{8pt}
|\rule[-2pt]{0.6pt}{12pt}
% \chapter*{Acknowledgements}
% \addcontentsline{toc}{chapter}{Acknowledgements}
% 
% 
% I hope that the people who helped me complete this project will appreciate how much they have helped me. This year has been an incredibly challenging journey---and perhaps I am not open enough about what it has been like. I undoubtedly would not have reached this day without everyone's help.
% 
% James Johnson---your mentorship and guidance throughout this project was critical. 
% You helped me learn an incredible amount about being an astronomer, and I am forever grateful for how you have guided this project, through late nights and last-minute comments.  Wayne, for everything you do for everyone---you are the glue holding together our department and nobody can say thank you enough. 
% David and Jennifer for helping this project move forward, by providing guidance at critical junctions and being part of my committee. Rolando, for being the non-astronomer thrown into this work, and pointing out the flaw of incomprehensibility in this work.
% 
% Even though you may have not been able to help me with the science, dear friends and family, you have been no less important in the completion of this work. 
% Eric, even across the country, you have always been a point of support and stability, and I am always happy to be around you. Anya, you are a wonderful friend and I cannot thank you enough for your unconditional support this year, 
% All my friends who are just wonderful, inspiring, and incredible people---Kaia, Maria, Autumn, Simon, Bailee, Aaliyah, Sanskruti, Alyssa, Harrison, Denis, Keith, and so many others. 
% And my family (including Arya, my most lovable doodle)---I can not even being to thank you enough for simply helping me recover my health and make it to graduation. 
% 
% 
%

@@

\chapter*{Symbols}

\setlength{\tabcolsep}{0pt}
\begin{longtable}{p{0.2\textwidth} p{0.8\textwidth}}
$\rho$ & 3-dimensional mass density. \\

$\Sigma$ & 2-dimensional density \\
\end{longtable}

\setlength{\tabcolsep}{6pt}
\renewcommand*{\arraystretch}{1}

\printglossary[type=\acronymtype,nonumberlist]
\printglossary

\newpage

@@

\the\textwidth

@@

>>>>

\title{From tides to tails: \\
    Does the Milky Way affect Sculptor and Ursa Minor?
}
\author{Daniel A. Boyea}

@@

\makeatletter
\begin{titlepage}
   \begin{center}
       \textbf{\large \@title}\\
        \skipline

@@

    in the Department of Physics and Astronomy\\
       \vfill
       {\small
       ©\@author, 2025\\
       University of Victoria\\
   }
   \skipline
       {\small
       All rights reserved. This thesis may not be reproduced in whole or in part,
   by photocopy or other means, without the permission of the author.} \\
\skipline
   \end{center}
We acknowledge and respect the \Lekwnen{} (Songhees and X\textsuperscript{w}seps\schwa{}m/
Esquimalt) Peoples on whose territory the university stands, and the
\Lekwnen{} and ${\underaccent{\bar}{\rm W}}$S\'ANE\'C Peoples whose historical relationships with the
land continue to this day. 
\end{titlepage}

@@

% Abstract of the paper
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
% context
\input{abstract.tex}

% \chapter*{Acknowledgements}
% \addcontentsline{toc}{chapter}{Acknowledgements}
%

@@

%\chapter*{Symbols}
%
%
%\setlength{\tabcolsep}{0pt}
%\begin{longtable}{p{0.2\textwidth} p{0.8\textwidth}}
%$\rho$ & 3-dimensional mass density. \\
%
%$\Sigma$ & 2-dimensional density \\
%\end{longtable}
%
%\setlength{\tabcolsep}{6pt}
%\renewcommand*{\arraystretch}{1}
%
%
%\printglossary[type=\acronymtype,nonumberlist]
%\printglossary
%
%
\newpage

@@

@@

\chapter{Font properties}

textwidth: \the\textwidth

-\rule[2.4pt]{3pt}{0.75pt}
I\rule{1.2pt}{8pt}
{\bf I}\rule{2.0pt}{8pt}
|\rule[-2pt]{0.6pt}{12pt}
